---
title: "Supplementary tables for “Impacts of competition and phenotypic plasticity on the viability of adaptive therapy”"
author:
  - B. Vibishan, Paras Jain^[Equal contribution], Vedant Sharma^[Equal contribution], Kishore Hari, Claus Kadelka,
  - Jason T. George^[Corresponding author], Mohit Kumar Jolly^[Corresponding author]
date: ""
output: 
  pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(psych)
library(ggplot2)
library(patchwork)
library(latex2exp)
library(lmerTest)
library(ggbiplot)
library(emmeans)
library(readxl)
library(broom)
library(readxl)
setwd('~/adaptive-therapy-with-transitions/analysis/')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

```{r data-processing, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
plotdf.cycling <- function(df, t_type){
  
  df$ModelType <- c(rep('AC+Tr', 10000), rep('AC', 10000), rep('SC+Tr', 10000), rep('SC', 10000)) 
  df <- df %>%
    mutate(flag = if_else((a > r_x)|(b > r_x)|(a > r_y)|((b> r_y)), TRUE, FALSE)) %>%
    filter(flag == FALSE) %>%
    select(-flag) %>%
    mutate(Outcome = if_else(Period != 'Inf', 'Cycling', if_else(ResFraction > 0.15, 'Unfavourable', 'Favourable')),
         Therapy = t_type,
         Outcome = factor(Outcome, levels = c('Unfavourable', 'Cycling', 'Favourable')))
  
  df_outcomes <- df %>%
    mutate(Outcome = if_else(Period != 'Inf', 'Cycling', if_else(ResFraction > 0.15, 'Unfavourable', 'Favourable')),
           Therapy = t_type) %>%
    group_by(ModelType, Therapy, Outcome) %>%
    count()
  
  df_outcomes <- within(df_outcomes, {ModelType = as.factor(ModelType)
                                              Therapy = as.factor(Therapy)
                                              Outcome = as.factor(Outcome)})
  df_outcomes <- df_outcomes %>%
    mutate(Outcome = factor(Outcome, levels = c('Unfavourable', 'Cycling', 'Favourable')))
  
  return(list("df" = df, "outcomes" = df_outcomes))
}

plotdf <- function(df, t_type){
  
  df$ModelType <- c(rep('AC+Tr', 10000), rep('AC', 10000), rep('SC+Tr', 10000), rep('SC', 10000)) 
  df <- df %>%
    mutate(flag = if_else((a > r_x)|(b > r_x)|(a > r_y)|((b> r_y)), TRUE, FALSE)) %>%
    filter(flag == FALSE) %>%
    select(-flag) %>%
    mutate(Outcome = if_else(ResFraction > 0.15, 'Unfavourable', 'Favourable'),
         Therapy = t_type,
         Outcome = factor(Outcome, levels = c('Unfavourable', 'Favourable')))
  
  df_outcomes <- df %>%
    mutate(Outcome = if_else(ResFraction > 0.15, 'Unfavourable', 'Favourable'),
           Therapy = t_type) %>%
    group_by(ModelType, Therapy, Outcome) %>%
    count()
  
  df_outcomes <- within(df_outcomes, {ModelType = as.factor(ModelType)
                                              Therapy = as.factor(Therapy)
                                              Outcome = as.factor(Outcome)})
  df_outcomes <- df_outcomes %>%
    mutate(Outcome = factor(Outcome, levels = c('Unfavourable', 'Favourable')))
  
  return(list("df" = df, "outcomes" = df_outcomes))
}
```

```{r no-treat-data-import, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE, cache=TRUE}
no_therapy_r1 <- read_csv("no-therapy-outcomes-r1.csv", 
    col_types = cols(...1 = col_skip()))
no_therapy_r2 <- read_csv("no-therapy-outcomes-r2.csv", 
    col_types = cols(...1 = col_skip()))
no_therapy_r3 <- read_csv("no-therapy-outcomes-r3.csv", 
    col_types = cols(...1 = col_skip()))

df <- plotdf(no_therapy_r1, 'No therapy')
no_therapy_r1 <- df$df
no_therapy_r1$RepNum <- 1
no_therapy_outcomes_r1 <- df$outcomes
no_therapy_outcomes_r1$RepNum <- 1

df <- plotdf(no_therapy_r2, 'No therapy')
no_therapy_r2 <- df$df
no_therapy_r2$RepNum <- 2
no_therapy_outcomes_r2 <- df$outcomes
no_therapy_outcomes_r2$RepNum <- 2

df <- plotdf(no_therapy_r3, 'No therapy')
no_therapy_r3 <- df$df
no_therapy_r3$RepNum <- 3
no_therapy_outcomes_r3 <- df$outcomes
no_therapy_outcomes_r3$RepNum <- 3

no_therapy <- bind_rows(no_therapy_r1, no_therapy_r2, no_therapy_r3)
no_therapy_outcomes <- bind_rows(no_therapy_outcomes_r1, no_therapy_outcomes_r2, no_therapy_outcomes_r3)

no_therapy_prop <- no_therapy_outcomes %>%
  pivot_wider(names_from = Outcome, values_from = n) %>%
  mutate(FavProp = Favourable / (Unfavourable + Favourable)) %>%
  group_by(ModelType, Therapy) %>%
  summarise(MeanFavProp = mean(FavProp),
            Propmin = MeanFavProp - sd(FavProp),
            Propmax = MeanFavProp + sd(FavProp)) %>%
  ungroup()

no_therapy_outcomes <- no_therapy_outcomes %>%
  pivot_wider(names_from = Outcome, values_from = n)
```

## Supplementary Table 1: Pairwise Wilcoxon's signed-rank tests comparing steady state population sizes across model types

```{r outcome-stats, message=FALSE, warning=FALSE}
popsize_df <- no_therapy %>%
  select(ModelType, PopSizeSS)

pairwise.wilcox.test(popsize_df$PopSizeSS, popsize_df$ModelType, p.adjust.method = 'BH', paired = F)
```

```{r cdt-data-import, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE, cache=TRUE}
st1_r1 <- read_csv("constant-dose/CDT-cytostatic-0.1-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))

df <- plotdf(st1_r1, 'Cytostatic1')
st1_r1 <- df$df
st1_r1$RepNum <- 1
st1_outcomes_r1 <- df$outcomes
st1_outcomes_r1$RepNum <- 1

st1_r2 <- read_csv("constant-dose/CDT-cytostatic-0.1-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(st1_r2, 'Cytostatic1')
st1_r2 <- df$df
st1_r2$RepNum <- 2
st1_outcomes_r2 <- df$outcomes
st1_outcomes_r2$RepNum <- 2

st1_r3 <- read_csv("constant-dose/CDT-cytostatic-0.1-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(st1_r3, 'Cytostatic1')
st1_r3 <- df$df
st1_r3$RepNum <- 3
st1_outcomes_r3 <- df$outcomes
st1_outcomes_r3$RepNum <- 3

st2_r1 <- read_csv("constant-dose/CDT-cytostatic-0.33-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(st2_r1, 'Cytostatic2')
st2_r1 <- df$df
st2_r1$RepNum <- 1
st2_outcomes_r1 <- df$outcomes
st2_outcomes_r1$RepNum <- 1

st2_r2 <- read_csv("constant-dose/CDT-cytostatic-0.33-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(st2_r2, 'Cytostatic2')
st2_r2 <- df$df
st2_r2$RepNum <- 2
st2_outcomes_r2 <- df$outcomes
st2_outcomes_r2$RepNum <- 2

st2_r3 <- read_csv("constant-dose/CDT-cytostatic-0.33-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(st2_r3, 'Cytostatic2')
st2_r3 <- df$df
st2_r3$RepNum <- 3
st2_outcomes_r3 <- df$outcomes
st2_outcomes_r3$RepNum <- 3

tx1_r1 <- read_csv("constant-dose/CDT-cytotoxic-0.01-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))

df <- plotdf(tx1_r1, 'Cytotoxic1')
tx1_r1 <- df$df
tx1_r1$RepNum <- 1
tx1_outcomes_r1 <- df$outcomes
tx1_outcomes_r1$RepNum <- 1

tx1_r2 <- read_csv("constant-dose/CDT-cytotoxic-0.01-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(tx1_r2, 'Cytotoxic1')
tx1_r2 <- df$df
tx1_r2$RepNum <- 2
tx1_outcomes_r2 <- df$outcomes
tx1_outcomes_r2$RepNum <- 2

tx1_r3 <- read_csv("constant-dose/CDT-cytotoxic-0.01-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(tx1_r3, 'Cytotoxic1')
tx1_r3 <- df$df
tx1_r3$RepNum <- 3
tx1_outcomes_r3 <- df$outcomes
tx1_outcomes_r3$RepNum <- 3

tx2_r1 <- read_csv("constant-dose/CDT-cytotoxic-0.05-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(tx2_r1, 'Cytotoxic2')
tx2_r1 <- df$df
tx2_r1$RepNum <- 1
tx2_outcomes_r1 <- df$outcomes
tx2_outcomes_r1$RepNum <- 1

tx2_r2 <- read_csv("constant-dose/CDT-cytotoxic-0.05-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(tx2_r2, 'Cytotoxic2')
tx2_r2 <- df$df
tx2_r2$RepNum <- 2
tx2_outcomes_r2 <- df$outcomes
tx2_outcomes_r2$RepNum <- 2

tx2_r3 <- read_csv("constant-dose/CDT-cytotoxic-0.05-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf(tx2_r3, 'Cytotoxic2')
tx2_r3 <- df$df
tx2_r3$RepNum <- 3
tx2_outcomes_r3 <- df$outcomes
tx2_outcomes_r3$RepNum <- 3

cdt_df <- bind_rows(st1_r1, st1_r2, st1_r3,
                    st2_r1, st2_r2, st2_r3,
                    tx1_r1, tx1_r2, tx1_r3,
                    tx2_r1, tx2_r2, tx2_r3)

cdt_df_outcomes <- bind_rows(st1_outcomes_r1, st1_outcomes_r2, st1_outcomes_r3, 
                             st2_outcomes_r1, st2_outcomes_r2, st2_outcomes_r3,
                             tx1_outcomes_r1, tx1_outcomes_r2, tx1_outcomes_r3,
                             tx2_outcomes_r1, tx2_outcomes_r2, tx2_outcomes_r3)

cdt_prop <- cdt_df_outcomes %>%
  pivot_wider(names_from = Outcome, values_from = n) %>%
  mutate(FavProp = Favourable / (Unfavourable + Favourable)) %>%
  group_by(ModelType, Therapy) %>%
  summarise(MeanFavProp = mean(FavProp),
            Propmin = MeanFavProp - sd(FavProp),
            Propmax = MeanFavProp + sd(FavProp)) %>%
  ungroup() %>%
  mutate(MeanFavProp = if_else(Therapy == 'Cytotoxic2', 0, MeanFavProp),
         Propmin = if_else(Therapy == 'Cytotoxic2', 0, Propmin),
         Propmax = if_else(Therapy == 'Cytotoxic2', 0, Propmax))

cdt_outcomes_mean <- cdt_df_outcomes %>%
  group_by(ModelType, Therapy, Outcome) %>%
  summarise(MeanCount = mean(n),
            SD = sd(n)) %>%
  ungroup()
```

## Supplementary Table 2: Mean differences in number of "Favourable" outcomes between dose levels under constant dose therapy (CDT)

```{r overall-favprop-stats, message=FALSE, warning=FALSE, echo=FALSE}
cdt_df_outcomes %>%
  pivot_wider(names_from = Outcome, values_from = n) %>%
  nest(data = c(-RepNum, -ModelType)) %>%
  mutate(fit = map(data, ~ lm(Favourable ~ Therapy, data = .)),
      mean.comp = map(fit, emmeans, ~Therapy, type = 'response'),
      odds.obj = map(mean.comp, pairs, reverse = T, simple = 'Therapy'),
      odds.ratio = map(odds.obj, tidy, conf.int = FALSE)) %>%
  unnest(odds.ratio) %>%
    select(RepNum, ModelType, contrast, estimate) %>%
    filter(contrast %in% c('Cytostatic2 - Cytostatic1', 'Cytotoxic2 - Cytotoxic1')) %>%
  group_by(ModelType, contrast) %>%
  summarise(MeanDiff= mean(estimate),
            Error = SD(estimate))
```

```{r adaptive-therapy-data-import, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE, cache=TRUE}
st1_r1 <- read_csv("adaptive-therapy/AT-cytostatic-0.1-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))

df <- plotdf.cycling(st1_r1, 'Cytostatic1')
st1_r1 <- df$df
st1_r1$RepNum <- 1
st1_outcomes_r1 <- df$outcomes
st1_outcomes_r1$RepNum <- 1

st1_r2 <- read_csv("adaptive-therapy/AT-cytostatic-0.1-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(st1_r2, 'Cytostatic1')
st1_r2 <- df$df
st1_r2$RepNum <- 2
st1_outcomes_r2 <- df$outcomes
st1_outcomes_r2$RepNum <- 2

st1_r3 <- read_csv("adaptive-therapy/AT-cytostatic-0.1-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(st1_r3, 'Cytostatic1')
st1_r3 <- df$df
st1_r3$RepNum <- 3
st1_outcomes_r3 <- df$outcomes
st1_outcomes_r3$RepNum <- 3

st2_r1 <- read_csv("adaptive-therapy/AT-cytostatic-0.33-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(st2_r1, 'Cytostatic2')
st2_r1 <- df$df
st2_r1$RepNum <- 1
st2_outcomes_r1 <- df$outcomes
st2_outcomes_r1$RepNum <- 1

st2_r2 <- read_csv("adaptive-therapy/AT-cytostatic-0.33-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(st2_r2, 'Cytostatic2')
st2_r2 <- df$df
st2_r2$RepNum <- 2
st2_outcomes_r2 <- df$outcomes
st2_outcomes_r2$RepNum <- 2

st2_r3 <- read_csv("adaptive-therapy/AT-cytostatic-0.33-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(st2_r3, 'Cytostatic2')
st2_r3 <- df$df
st2_r3$RepNum <- 3
st2_outcomes_r3 <- df$outcomes
st2_outcomes_r3$RepNum <- 3

tx1_r1 <- read_csv("adaptive-therapy/AT-cytotoxic-0.01-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))

df <- plotdf.cycling(tx1_r1, 'Cytotoxic1')
tx1_r1 <- df$df
tx1_r1$RepNum <- 1
tx1_outcomes_r1 <- df$outcomes
tx1_outcomes_r1$RepNum <- 1

tx1_r2 <- read_csv("adaptive-therapy/AT-cytotoxic-0.01-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(tx1_r2, 'Cytotoxic1')
tx1_r2 <- df$df
tx1_r2$RepNum <- 2
tx1_outcomes_r2 <- df$outcomes
tx1_outcomes_r2$RepNum <- 2

tx1_r3 <- read_csv("adaptive-therapy/AT-cytotoxic-0.01-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(tx1_r3, 'Cytotoxic1')
tx1_r3 <- df$df
tx1_r3$RepNum <- 3
tx1_outcomes_r3 <- df$outcomes
tx1_outcomes_r3$RepNum <- 3

tx2_r1 <- read_csv("adaptive-therapy/AT-cytotoxic-0.05-zero-delay-r1.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(tx2_r1, 'Cytotoxic2')
tx2_r1 <- df$df
tx2_r1$RepNum <- 1
tx2_outcomes_r1 <- df$outcomes
tx2_outcomes_r1$RepNum <- 1

tx2_r2 <- read_csv("adaptive-therapy/AT-cytotoxic-0.05-zero-delay-r2.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(tx2_r2, 'Cytotoxic2')
tx2_r2 <- df$df
tx2_r2$RepNum <- 2
tx2_outcomes_r2 <- df$outcomes
tx2_outcomes_r2$RepNum <- 2

tx2_r3 <- read_csv("adaptive-therapy/AT-cytotoxic-0.05-zero-delay-r3.csv", 
    col_types = cols(...1 = col_skip()))
df <- plotdf.cycling(tx2_r3, 'Cytotoxic2')
tx2_r3 <- df$df
tx2_r3$RepNum <- 3
tx2_outcomes_r3 <- df$outcomes
tx2_outcomes_r3$RepNum <- 3

at_df <- bind_rows(st1_r1, st1_r2, st1_r3,
                    st2_r1, st2_r2, st2_r3,
                    tx1_r1, tx1_r2, tx1_r3,
                    tx2_r1, tx2_r2, tx2_r3)

at_df_outcomes <- bind_rows(st1_outcomes_r1, st1_outcomes_r2, st1_outcomes_r3, 
                             st2_outcomes_r1, st2_outcomes_r2, st2_outcomes_r3,
                             tx1_outcomes_r1, tx1_outcomes_r2, tx1_outcomes_r3,
                             tx2_outcomes_r1, tx2_outcomes_r2, tx2_outcomes_r3)
```

## Supplementary Table 3: Pairwise Wilcoxon's signed-rank tests comparing cycling time periods across model types
```{r cycling-stats, message=FALSE, warning=FALSE, cache=TRUE}
cycling_df <- at_df %>%
  filter(Outcome == 'Cycling') %>%
  select(ModelType, Period)

pairwise.wilcox.test(cycling_df$Period, cycling_df$ModelType, p.adjust.method = 'BH', paired = F)
```

```{r import-analyses, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE, cache=TRUE}
vc.r1 <- read_csv('adaptive-therapy/virtual-cohort-cytotoxic-0.05-r1.csv',
         col_types = cols(...1 = col_skip()))
vc.r1$RepNum <- 1

vc.r2 <- read_csv('adaptive-therapy/virtual-cohort-cytotoxic-0.05-r2.csv',
         col_types = cols(...1 = col_skip()))
vc.r2$RepNum <- 2

vc.r3 <- read_csv('adaptive-therapy/virtual-cohort-cytotoxic-0.05-r3.csv',
         col_types = cols(...1 = col_skip()))
vc.r3$RepNum <- 3

vc.all <- bind_rows(vc.r1, vc.r2, vc.r3) %>%
  select(-t_delay) %>%
  mutate(ModelType = if_else(c==d, if_else(a==b, 'SC', 'SC+Tr'), if_else(a==b, 'AC', 'AC+Tr')))

vc_static <- vc.all %>%
  select(-ResFracAT, -PeriodAT) %>%
  pivot_longer(cols = c(ResFracBT, ResFracCDT), names_to = 'Therapy', values_to = 'ResFrac', names_prefix = 'ResFrac') %>%
  mutate(Outcome = if_else(ResFrac > 0.15, 'Unfavourable', 'Favourable'))

vc_static_outcomes <- vc_static %>%
  group_by(ModelType, Therapy, Outcome, RepNum) %>%
  count() %>% ungroup()


vc_at <- vc.all %>%
  select(-ResFracBT, -ResFracCDT) %>%
  pivot_longer(cols = c(ResFracAT, PeriodAT), names_to = c('Param', 'Therapy'), values_to = 'value', names_pattern = '([A-Za-z]+)([A-Za-z]+[A-Za-z]+)') %>%
  pivot_wider(names_from = Param, values_from = value) %>%
  mutate(Outcome = if_else(Period != 'Inf', 'Cycling', if_else(ResFrac > 0.15, 'Unfavourable', 'Favourable')))

vc_at_outcomes <- vc_at %>%
  group_by(ModelType, Therapy, Outcome, RepNum) %>%
  count() %>% ungroup()

vc_all_outcomes <- bind_rows(vc_static_outcomes, vc_at_outcomes) %>%
  mutate(Outcome = factor(Outcome, levels = c('Unfavourable', 'Favourable', 'Cycling')),
         Therapy = factor(Therapy, levels = c('BT', 'CDT', 'AT')))

vc_params_long <- bind_rows(vc_static, vc_at) %>%
  pivot_longer(cols = r_x:d, names_to = 'Parameter', values_to = 'value') %>%
  mutate(Outcome = factor(Outcome, levels = c('Unfavourable', 'Favourable', 'Cycling')),
         Therapy = factor(Therapy, levels = c('BT', 'CDT', 'AT')),
         Parameter = factor(Parameter, levels = c('r_x', 'r_y', 'a', 'b', 'c', 'd')))

cdt_pca_df <- vc.all %>%
  select(-ResFracAT, -PopSizeAT, -PeriodAT) %>%
  mutate(Outcome = if_else(ResFracCDT > 0.15, 'Unfavourable', 'Favourable'))

at_pca_df <- vc.all %>%
  select(-ResFracCDT, -PopSizeCDT) %>%
  mutate(Outcome = if_else(PeriodAT != 'Inf', 'Cycling', if_else(ResFracAT > 0.15, 'Unfavourable', 'Favourable')))
```

## Supplementary Table 4: PC loadings and component variances for resistant cell fraction before and after constant dose therapy
```{r cdt-resfraction, cache=TRUE}
pca.cdt <- prcomp(~ r_x + r_y + c + d + a + b + ResFracBT + ResFracCDT, data = cdt_pca_df, scale. =T)
pca.cdt
summary(pca.cdt)
```

## Supplementary Table 5: PC loadings and component variances for total population size before and after constant dose therapy
```{r cdt-popsize, cache=TRUE}
pca.cdt <- prcomp(~ r_x + r_y + c + d + a + b + PopSizeBT + PopSizeCDT, data = cdt_pca_df, scale. =T)
pca.cdt
summary(pca.cdt)
```

## Supplementary Table 6: PC loadings and component variances for resistant cell fraction before and after adaptive therapy
```{r at-resfraction, cache=TRUE}
pca.at <- prcomp(~ r_x + r_y + c + d + a + b + ResFracBT + ResFracAT, data = at_pca_df, scale. =T)
pca.at
summary(pca.at)
```

## Supplementary Table 7: PC loadings and component variances for total population size before and after adaptive therapy
```{r at-popsize, cache=TRUE}
pca.at <- prcomp(~ r_x + r_y + c + d + a + b + PopSizeBT + PopSizeAT, data = at_pca_df, scale. =T)
pca.at
summary(pca.at)
```